---
title: "Data Analysis Assignment"
author: "Sam Johnson"
date: "2024-03-15"
format: html
---

# Introduction

This is my assignment for R class. I will try to answer the questions.

## Question 1: Understanding R

**Explain what R is and describe its main uses in data science:**

R is a computer program for math stuff. Its used by people who work with numbers and data. I think its free to download.

You can use R to:
- Do math
- Make graphs
- Look at data

I'm not sure about all the details but I know it's popular.

## Question 2: Basic Programming

**Write code to calculate descriptive statistics for a numeric vector:**

```{r stats}
# Make some numbers
numbers <- c(1, 2, 3, 4, 5)

# Calculate mean
mean(numbers)

# Other stuff
median(numbers)
```

## Question 3: Data Visualization

**Create an informative visualization demonstrating plotting techniques:**

```{r plot}
# Make a simple plot
x <- c(1, 2, 3, 4, 5)
y <- c(2, 4, 6, 8, 10)

plot(x, y)
```

```{r another-plot}
# Try to make another plot but I'm not sure how
# library(ggplot2)
# This doesn't work for some reason
# ggplot(data, aes(x, y)) + geom_point()

# Just use basic plot instead
plot(1:10, (1:10)^2, type = "l")
```

## Question 4: More stuff

**Do some analysis:**

```{r analysis}
# I don't really know what else to do
data1 <- c(1, 3, 5, 7, 9)
data2 <- c(2, 4, 6, 8, 10)

# Maybe correlation?
# cor(data1, data2)  # This might work

# Just print the data
print(data1)
print(data2)

# Try histogram
hist(data1)
```

# Conclusion

I did the assignment and answered all the questions. R seems complicated but I tried my best. 

The code mostly works I think. I had trouble with some parts but I included what I could figure out. Maybe next time I can do better if I study more.

I know R is important for data science but I need more practice to understand it better.